The Civil Society Organizations effect: A mixed-Methods analysis of bottom-up approaches in Brazilian public policy

### 2023-11-02 Scripts/paper.Rmd
Referências bib etc.

---
title: "Unlocking Accountability: A Mixed-Methods Analysis of Bottom-Up Approaches in Brazilian Public Policy"
date: "2023-10-26"
output: pdf_document
sansfont: Sans Serif
fontsize: 12pt
author:
- name: Manoel Galdino
  affiliation: Universidade de São Paulo
abstract: "This paper examines the role of accountability in democratic systems, focusing on its impact on public service delivery. Employing mixed methods, we differentiate between information-driven interventions and the mobilization and monitoring efforts of organized Civil Society Organizations (CSOs). We argue that well-organized CSOs, deeply embedded in society and politics, can drive significant policy change. The study evaluates the effectiveness of the Obra Transparente project by Transparência Brasil, engaging 21 local CSOs in South and Southeast Brazil. In contrast to information-centered approaches, this project establishes a network of local CSOs, providing technical support and training for monitoring public school and nursery construction projects. Our findings emphasize the importance of sustained, coordinated efforts by organized civil society. These efforts, involving direct engagement with municipal officials, persistent follow-ups, and on-site visits, lead to more substantial outcomes. This research contributes to understanding how organized civil society mobilization can achieve bottom-up accountability, a crucial aspect of democratic governance. However, contextual variations may limit the effectiveness of CSO interventions."
keywords: "Accountability, Brazil, mixed methods, state capacity, public policy"
---

@@

```{r package loads, message=F, echo=FALSE}
library(data.table)
library(here)
library(janitor)
library(lubridate)
library(readxl)
library(tidyverse)
library(gt)
library(ipeadatar)
library(MatchIt)
library(marginaleffects)

@@

# lista_cidades_ot <- c("Araucária", "Caçador", "Campo Mourão", "Cascavel", "Chapecó",
#                       "Foz do Iguaçu", "Goioerê", "Gravataí", "Guarapuava", "Imbituba", "Lajeado",
#                       "Limeira", "Palhoça", "Paranaguá", "Pelotas", "Ponta Grossa", "Santa Maria",
#                       "São Francisco do Sul", "São José dos Campos", "Taubaté", "Uberlândia")
# 
# df_cidades_ot <- data.frame(municipio = lista_cidades_ot, indicator_ot = 1)

@@

df_cidades_ot <- read_excel(here("Dados", "181114 - Situação obras_atual.xlsx"), 
                                           sheet = "SIMEC-Mai17", skip = 1) %>%
  clean_names() %>%
  filter(!is.na(id)) %>%
  dplyr::select(id, municipio, uf, situacao2, levant_os_jun_17) %>%
  mutate(indicator_ot = 1)
```

@@

antes_projeto <- obras_inicio_projeto_ot %>%
  group_by(indicator_muni_ot, concluida) %>%
  summarise(contagem = n()) %>%
  mutate(total = sum(contagem),
         perc = round(contagem/total, 2))

@@

depois_projeto <- obras_fim_seg_fase %>%
  group_by(indicator_muni_ot, concluida) %>%
  summarise(contagem = n()) %>%
  mutate(total = sum(contagem),
         perc = round(contagem/total, 2))

@@

Accountability is a vital component of democratic systems to ensure the efficient delivery of public services (Besley and Ghatak, 2003; Cameron, 2004;  O'Donnell, 1998). Traditionally defined as the process of holding authorities responsible for their actions (Finer, 1941; Mulgan, 2000; O'Loughlin, 1990), the literature has evolved to encompass a broader range of mechanisms for overseeing politicians and the bureaucracy.. While some authors focused on the study of formal institutions (O’Donnel, 2003, Kenney 2003) in what is known as horizontal accountability, other have analyzed responsiveness of politicians to voters (Relly, 2012) the media and civil society organizations at large (Fox e Brown, 1998; Paul, 2002); in what is known as vertical accountability.

Recent studies have highlighted the role of "bottom-up accountability" (also known as social accountability, Fox, 2015), in which citizens receive information about public policies and use it for social control. The pioneering study 'Power to the People' (Bjorkman and Svensson, 2009), which demonstrated the impact of local community engagement in monitoring the provision of health services, inspired subsequent literature that tested similar interventions in different contexts.

The mechanism would works like this: Constituents have firsthand information about the outcomes of local policies, giving citizens incentives to combat corruption that directly affects them. Since policymakers are sensitive to social feedback from their own communities (Serra, 2011: 570), bottom-up accountability could influence local public policies

However, a series of recent studies have raised doubts about the effectiveness of "bottom-up accountability" strategies based on interventions aimed at increasing the monitoring capacity of ordinary citizens regarding public policies (Freire, Mignozzetti & Galdino, 2020; Raffler, Posner & Parkinson, 2019, Fox, 2015).

@@

The disappointing results have cast doubts on the effectiveness of the bottom-up accountability strategy for improving the delivery of public services, reducing corruption, and enhancing government efficiency. Some authors (Fox, 2015; Ankamah, 2019) have suggested that social accountability without "teeth," defined as the government's capacity to respond to citizens' voices, has limited impact.

@@

We argue that we need to differentiate between information-led tactical interventions and active mobilization and monitoring by Civil Society Organizations (CSOs). By their very nature, established CSOs are embedded in society and politics and can have a substantial impact on policy change (Gong & Xiao, 2017. Joshi & Houtzager, 2012).

The paper argues that a missing feature in most of the literature is that these interventions should target CSOs instead of individual citizens, and ensure that the intervention enhances the ability of the CSOs to overcome their collective action problem  (Mattoni & Odilla, 2021; Odilla, 2023).

The present paper provides evidence supporting this argument. We reassess the evidence provided by Freire, Mignozzetti & Galdino (2020), considering a different intervention than the one they evaluated. They provided experimental evidence that a mobile app (Tá de Pé) made available to citizens did not improve the delivery of school and nursery construction works. Here, we assess the efficacy of a sister project, called "Obra Transparente," also developed by the Brazilian NGO Transparência Brasil. This project targeted 21 local Civil Society Organizations in 21 municipalities in the South and Southeast regions of Brazil, with the aim of improving access to education by addressing mismanagement and delays in the construction of public schools and nurseries. Both projects were developed in parallel, with similar goals but achieved strikingly different results.

The argument in the present study is that not all types of interventions that promote citizens' voice are the same. We argue that we need to differentiate between information-led tactical interventions and active mobilization and monitoring by Civil Society Organizations. By their very nature, established CSOs are embedded in society and politics and, as such, can have teeth of their own (Gong & Xiao, 2017. Joshi & Houtzager, 2012).  It may be only that we need mobilization by organized civil society, in a structured way, for civil society voices to be heard with “teeth”. Some studies have suggested that a key feature of successful change in public policies is the pressure of collectively organized agents (Mattoni & Odilla, 2021; Odilla, 2023). Collective action is a hard problem to be overcome and creating CSOs is one way to achieve this. We argue in this paper that a missing features of the tactical interventions assessed in most of the literature is that they should target CSOs instead of individual citizens, and make sure that the intervention increases the ability of the CSOs to attract and overcome their collective action problem.

@@

We employ an integrated inference approach (Humphreys & Jacobs, 2023) to analyze qualitative data within the framework of causal queries. As this approach is relatively innovative within qualitative research, we'll provide a comprehensive explanation.

The methodology outlined by Humphreys & Jacobs (2023) mandates the creation of a causal model encoded in Directed Acyclic Graphs (DAGs), a framework developed by Pearl (2009). Bayesian inference, recently promoted in qualitative data analysis (Fairfield & Charman, 2022), serves as the foundation for deriving quantitative estimates for causal queries constructed using the proposed DAG. 
In our current study, we investigate the impact of Civil Society Organization (CSO) monitoring on two aspects: the completion of construction works and the rectification of issues and irregularities found in construction facilities. Formally, the first research question relates to a causal relationship between monitoring (M) and completion (C), indicated by the DAG (M -> C). The second research query delves into whether monitoring (M) causes the fixing of issues (F), or the (DAG M -> F). To accommodate the qualitative nature of our variables, we treat them as binary.

@@

This approach hinges on the notion that each case can belong to a specific type, denoting different causal relationships. In our current investigation, let's focus on the (M -> C) DAG. To clarify what a type is, we'll employ the Potential Outcomes (PO) notation (Rubin, 2005; Imbens & Rubin, 2015). Within the PO framework, every case has two potential outcomes: one when the treatment (in this case, CSO monitoring, M) is absent and another when the treatment is present. In simplified notation, if a city only completes construction works in the absence of monitoring, we express this in PO notation as C(1) = 0. If there is no monitoring (M = 0) and no completion, then we write C(0) = 0. Conversely, if there is monitoring (M = 1) and completion, we write C(1) = 1, and C(0) = 1 if there is completion without monitoring.

@@

The advantage of this approach is that all assumptions and restrictions imposed to identify the parameters of interest are transparent. It also allows the user to update posterior probabilities based on qualitative data. One reason for this is that with Bayesian inference, we do not need to rely on asymptotic theory to make inferences. Thus, there is no need for a large-n study. Formally speaking, one can make causal queries without any data, based only on prior probabilities for the parameters. Obviously the answer to such causal queries will be vague enough to be almost useless. But the point is that, once you have a prior and a causal model, encoded in a DAG, you can answer causal queries. As one collects data (any data, be it qualitative or quantitative), one can make assumptions about the nature of the data (such as the Data Generating Process), and then update the model and provide new answers to the causal queries, which hopefully will be informative enough to be useful for the task at hand. And even when the model or hypothesis is not completely identified, it can be partially identified (Manski, 2003), which is still a useful answer to causal queries.

@@

The evidence gathered is primarily based on documented reports collected by Transparência Brasil over the course of the project. We also use evidence collected on semi-structured interviews and an online survey with representatives of the 21 CSOs conducted by an independent evaluator of the project. We also use administrative data about construction works data by municipalities and socio-demographic data on municipalities. The questions asked in the survey as well the aggregate answered are presented in annex A.

@@

To put in perspective what was accomplished regarding the completion of all four projects, by the end of the project, of all 135 construction works planned for all 21 municipalities, only 25 were effectively delivered, a 18.5% delivery rate. In Taubaté,  before the monitoring by the local CSO, the delivery rate was zero. After that, the delivery rate was 100% and one of the construction work was a previously paralyzed construction. According to a report by Transparência Brasil (Coelho, Galdino & Sakai., 2021), out of 771 stalled construction projects that were eventually completed, over 50% of them took more than two years to be delivered and 25% took three years or more. As we can see, the odds were against finishing the construction works.

@@

From a causal perspective, we are interested in what is called case-level causal attribution (Humphreys & Jacobs, 2023), i.e, did the monitoring cause the delivery of the construction works?In other words, we are conditioning on the fact that M = 1 and C = 1, which constrains the possible types for Taubaté to be $\theta^C_{1,1}$ ou $\theta^C_{0,1}$. Assuming a simple model in which CSO monitoring causes the completion of works (M -> C), and with the eight within-case instances, we do the causal query. Our stimand of interest is the probability that Taubaté is a type $\theta^D_{10}$.

@@

In 2017, the Brazilian municipality of Araucária conducted bidding processes seeking to hire companies to build three nurseries. The project sent technical experts to the 3 planned construction sites, together with volunteers from the local Observatório, to check whether plans were consistent with the physical conditions observed on site. The experts’ assessment was that high-cost contention walls included in the construction plans were unnecessary or could be replaced by low-cost solutions through adjustments. Originally, the contention walls would cost R$ 1,577,338.57 (US$ 419,504.9, at the nominal exchange rate in 2019). The analysis was first directed to the local administration, suggesting changes to the construction plans, which were rejected. The bidding processes went on as initially planned. Subsequently, the experts’ findings were submitted to the Brazilian Supreme Audit Institution, Tribunal de Contas da União (TCU), by Transparência Brasil and Observatório Social do Brazil (see annex 1). TCU recommended that the bidding process should be redone, and construction plans were to be redesigned, with more cost-effective solutions for the contention walls. In the new bidding process, their cost plummeted to R$ 416,883.17 (US$ 110,873.2), a reduction of US$ 308,631.7 — 74% less. To give a sense of the impact on the city budget, the amount saved on these three constructions alone represents about 3% of all capital investments of Araucária in 2018. In short, in a single municipality, the project resulted in savings higher than the cost of the entire project funded by the grantee UNDEF (US$ 220,000.00).

@@

To do so, we collected data on school and nursery construction projects across the entire country, encompassing over two thousand municipalities and more than 14 thousand construction projects, spanning all phases from planning to execution and completion. The table below provides descriptive statistics for these construction projects.

@@

```{r tables, echo=FALSE, message=FALSE, results='asis'}
antes_projeto %>%
  inner_join(depois_projeto, by = "indicator_muni_ot") %>%
  gt() %>%
  tab_header(
    title = "Percentage of finished fonstruction works before the project",
    subtitle = "Situation in August 2017 and August 2019") %>%
  cols_label(
    indicator_muni_ot = "Participating Municipalities in Obra Transparente ",
    perc.x = "Percentage of Completed Constructions Before Project",
    perc.y = "Percentage of Completed Constructions After Project"
  ) %>%
    fmt_percent(
    columns = c(perc.x, perc.y),
    decimals = 0
  )
```

Our primary objective is to compare the project's effects on the rate of delivery. To estimate the Average Treatment Effect on the Treated (ATT), we employed a matching strategy (Ho et al. (2007; Rubin, 1973). We also applied the same procedure to pre-project data to ensure that there were no substantial differences between the treated and control groups.

We estimate all models using the a varying-intercept regression equation (Gelman & Hill, 2006):

$$
Y_i = \alpha + \beta_1 T_i + \gamma X_i + \theta Z_i + \epsilon_i
$$
where $i$ indexes the construction units. $Y_i$ is $1$ if a construction work is finished, and $0$ otherwise, α is the intercept, β denotes the average treatment effect on the treated (ATT), and $T_i$ is a binary treatment indicator (municipality is part or not of Obra Transparent). $\gamma$ is a vector of varying-intercept, $X_i$ is a matrix of Brazilian municipalities random effects, $\theta$ is a vector of controls, and $Z_i$ an array of control variables for the case $i$. The error term is denoted by $ε_i$. We fit a Bayesian model with rstanarm (Goodrich, 2023) with default weakly informative priors on parameters (Gelman et. al., 2008).

@@

```{r prep data matching,  message=FALSE, echo=FALSE}

@@

@@

Table (ref number) summarizes the main results of the regression.

# these are the no redundancy covariates
covs.nr <- c('perc_pea18', 'ext_pobres', 'log_rpc', 'p10_ricos_desigualdade', 'freq_escola_15_1', 'analfabetismo',
             'log_renda', 'log_pop', 'percentual_de_execucao', 'tipo_do_projeto')

ps_spec <- formula(indicator_muni_ot ~ log_rpc  + ext_pobres + freq_escola_15_1 + analfabetismo +
                     log_renda   +  p10_ricos_desigualdade +
                       + percentual_de_execucao)

set.seed(1234)
ps_fit_pos <- stan_glm(ps_spec, family=binomial(link='logit'), 
                     data=df_balancing_aux_post_se, algorithm='optimizing', refresh=0, iter = 2000)
# extrqacting logit pscores
pscores_pre <- apply(posterior_linpred(ps_fit_pre), 2, mean)

# matching
#matching without replacement, original formula

matches_pre <- matching(z=df_balancing_aux_se$indicator_muni_ot, score=pscores_pre, replace=FALSE)
matched_pre <- df_balancing_aux_se[matches$match.ind,]

matches_pos <- matching(z=df_balancing_aux_post_se$indicator_muni_ot, score=pscores_pos, replace=FALSE)
matched_pos <- df_balancing_aux_post_se[matches$match.ind,]

@@

```{r fitting bayesian model, message=FALSE, echo=FALSE}

te_spec_nr <- update(ps_spec, concluida_num ~ indicator_muni_ot + (1 | municipio) + .)

set.seed(20)
reg_ps_pre <- stan_lmer(te_spec_nr, data=df_balancing_aux_se[matches_pre$match.ind,])

## After
set.seed(20)
reg_ps_pos <- stan_lmer(te_spec_nr, data=df_balancing_aux_post_se[matches_pos$match.ind,])

summary(reg_ps_pre)['indicator_muni_ot', 1:5]
summary(reg_ps_pos)['indicator_muni_ot', 1:5]

```{r table 2, echo=FALSE,  message=FALSE, results='asis'}
# posterior draws

@@

# Combine summary statistics of posterior simulation draws
b1_df_after_cred <- data.frame(variables = "ATT After project",
                             post_mean = b1_mean_after,
                             post_sd = b1_sd_after,
                             b1_quant_after[2], 
                             b1_quant_after[1],
                             b1_quant_after[3])
b1_df <- bind_rows(b1_df_pre_cred, b1_df_after_cred)

b1_df %>% gt() %>%
  tab_header(
    title = "Effect of local OSB social monitoring on construction delivery rates") %>%
  cols_label(
    variables = "Variables",
    post_mean = "Posterior mean",
    post_sd = "Posterior sd",
    Q50 = "Posterior median",
    Q2.5 = "Posterior 2.5% CI",
    Q97.5 = "Posterior 97.5% CI"
  ) %>%
  fmt_number(
    columns = c(post_mean, post_sd, Q50, Q2.5, Q97.5 ),
    decimals = 3,
    use_seps = FALSE
  ) %>%
  tab_source_note(
    source_note = "452 observations after matching, with municipalities of same states as participants of Obra Transparente."
  ) 
```

@@

Ankamah, S. S. (2019). Why do “teeth” need “voice”? The case of anti‐corruption agencies in three Australian states. Australian Journal of Public Administration, 78(4), 481-496.

Besley T and Ghatak M (2003) Incentives, choice, and accountability
in the provision of public services. Oxford Review of Economic Policy 19(2): 235–249.

Björkman, M., & Svensson, J. (2009). Power to the people: evidence from a randomized field experiment on community-based monitoring in Uganda. The Quarterly Journal of Economics, 124(2), 735-769.

>>>>

---
title: "The Civil Society Organizations effect: A mixed-Methods analysis of bottom-up approaches in Brazilian public policy"
date: "2023-10-26"
output: bookdown::pdf_document2
sansfont: Sans Serif
fontsize: 12pt
author:
  - name: Manoel Galdino
    affiliation: Universidade de São Paulo - Brazil.
  - names: Juliana Sakai
    affiliation: Transparência Brasil - Brazil
  - names: Bianca Vaz Mondo
    affiliation: Transparência Brasil - Brazil
  - names: Natália Paiva
    affiliation: Alandar Consulting - Brazil
abstract: "This paper examines the effect of bottom-up accountability on public service delivery. We differentiate between information-driven interventions and the mobilization and monitoring efforts of organized Civil Society Organizations (CSOs), and argue that the latter type of interventions can can drive significant policy change. The study evaluates the effectiveness of the Obra Transparente project by Brazilian NGO Transparência Brasil, engaging 21 local CSOs in South and Southeast Brazil.  Our findings emphasize the importance of sustained, coordinated efforts by socially embedded CSOS. These efforts, involving direct engagement with municipal officials and ensuring that their complains cannot be ignored lead to more substantial outcomes when compared with citizens information-driven."
keywords: "accountability, Civil Society Organizations, mixed methods, community-driven development, public policy."
bibliography: OT_paper.bib
---

@@

```{r package loads, message=F, echo=FALSE}
library(data.table)
library(here)
library(janitor)
library(lubridate)
library(readxl)
library(tidyverse)
library(gt)
library(ipeadatar)
library(rstan)
library(arm)
library(rstanarm)
library(survey)
library(broom)
library(broom.mixed)
library(MetBrewer)
```

@@

@@

df_cidades_ot <- read_excel(here("Dados", "situacao obras_atual.xlsx"), 
                                           sheet = "SIMEC-Mai17", skip = 1) %>%
  clean_names() %>%
  filter(!is.na(id)) %>%
  dplyr::select(id, municipio, uf, situacao2, levant_os_jun_17) %>%
  mutate(indicator_ot = 1)
```

@@

antes_projeto <- obras_inicio_projeto_ot %>%
  dplyr::filter(uf %in% c("SP", "PR", "SP", "SC", "RS", "MG")) %>%
  group_by(indicator_muni_ot, concluida) %>%
  summarise(contagem = n()) %>%
  mutate(total = sum(contagem),
         perc = round(contagem/total, 2))

@@

depois_projeto <- obras_fim_seg_fase %>%
  dplyr::filter(uf %in% c("SP", "PR", "SP", "SC", "RS", "MG")) %>%
  group_by(indicator_muni_ot, concluida) %>%
  summarise(contagem = n()) %>%
  mutate(total = sum(contagem),
         perc = round(contagem/total, 2))

@@

Accountability is a crucial element in democratic systems, ensuring the efficient delivery of public services [@besley_2003; @cameron_2004; @odonnell_1998]. Initially defined as the process of holding authorities responsible for their actions [@mulgan_2000; @oloughlin_1990], the literature has since broadened to include various oversight mechanisms for politicians and the bureaucracy. Some authors focus on formal institutions [@odonnell_2003; @kenney_2003], known as horizontal accountability, while others examine how politicians respond to voters, the media, and civil society organizations [@bertelli_2020; @relly_2012; @paul_1992], referred to as vertical accountability.

Recent studies have highlighted the role of "bottom-up accountability" (also known as social accountability, Fox, 2015), in which citizens receive information about public policies and use it for social control. The pioneering study 'Power to the People' [@bjorkman_2009], which demonstrated the impact of local community engagement in monitoring the provision of health services, inspired subsequent literature that tested similar interventions in different contexts.

The mechanism would works like this: Constituents have firsthand information about the outcomes of local policies, giving citizens incentives to combat corruption that directly affects them. Since policymakers are sensitive to social feedback from their own communities [@serra_2012], bottom-up accountability could influence local public policies

However, a series of recent studies have raised doubts about the effectiveness of "bottom-up accountability" strategies based on interventions aimed at increasing the monitoring capacity of ordinary citizens regarding public policies [@freire_etal2020; @raffler_etal2019; @fox_2015].

@@

The disappointing results have cast doubts on the effectiveness of the bottom-up accountability strategy for improving the delivery of public services, reducing corruption, and enhancing government efficiency. Some authors [@fox_2015; @ankamah_2019] have suggested that social accountability without "teeth," defined as the government's capacity to respond to citizens' voices, has limited impact.

@@

We argue that we need to differentiate between information-led tactical interventions and active mobilization and monitoring by Civil Society Organizations (CSOs). By their very nature, established CSOs are embedded in society and politics and can have a substantial impact on policy change [@gong_2017; @joshi_2012].

The paper argues that a missing feature in most of the literature is that these interventions should target CSOs instead of individual citizens, and ensure that the intervention enhances the ability of the CSOs to overcome their collective action problem [@mattoni_2021; @odilla_2023].

The present paper provides evidence supporting this argument. We reassess the evidence provided by @freire_etal2020, considering a different intervention than the one they evaluated. They provided experimental evidence that a mobile app (Tá de Pé) made available to citizens did not improve the delivery of school and nursery construction works. Here, we assess the efficacy of a sister project, called "Obra Transparente," also developed by the Brazilian NGO Transparência Brasil. This project targeted 21 local Civil Society Organizations in 21 municipalities in the South and Southeast regions of Brazil, with the aim of improving access to education by addressing mismanagement and delays in the construction of public schools and nurseries. Both projects were developed in parallel, with similar goals but achieved strikingly different results.

The argument in the present study is that not all types of interventions that promote citizens' voice are the same. We argue that we need to differentiate between information-led tactical interventions and active mobilization and monitoring by Civil Society Organizations. By their very nature, established CSOs are embedded in society and politics and, as such, can have teeth of their own [@gong_2017; @joshi_2012].  It may be only that we need mobilization by organized civil society, in a structured way, for civil society voices to be heard with “teeth”. Some studies have suggested that a key feature of successful change in public policies is the pressure of collectively organized agents [@mattoni_2021; @odilla_2023]. Collective action is a hard problem to be overcome and creating CSOs is one way to achieve this. We argue in this paper that a missing features of the tactical interventions assessed in most of the literature is that they should target CSOs instead of individual citizens, and make sure that the intervention increases the ability of the CSOs to attract and overcome their collective action problem.

@@

We employ an integrated inference approach [@humphreys_2023] to analyze qualitative data within the framework of causal queries. As this approach is relatively innovative within qualitative research, we'll provide a comprehensive explanation.

The methodology outlined by [@humphreys_2023] mandates the creation of a causal model encoded in Directed Acyclic Graphs (DAGs), a framework developed by Pearl (2009). Bayesian inference, recently promoted in qualitative data analysis [@fairfield_2022], serves as the foundation for deriving quantitative estimates for causal queries constructed using the proposed DAG.

In our current study, we investigate the impact of Civil Society Organization (CSO) monitoring on two aspects: the completion of construction works and the rectification of issues and irregularities found in construction facilities. Formally, the first research question relates to a causal relationship between monitoring (M) and completion (C), indicated by the DAG (M -> C). The second research query delves into whether monitoring (M) causes the fixing of issues (F), or the (DAG M -> F). To accommodate the qualitative nature of our variables, we treat them as binary.

@@

This approach hinges on the notion that each case can belong to a specific type, denoting different causal relationships. In our current investigation, let's focus on the (M -> C) DAG. To clarify what a type is, we'll employ the Potential Outcomes (PO) notation [@rubin_2005; @imbens_2015]. Within the PO framework, every case has two potential outcomes: one when the treatment (in this case, CSO monitoring, M) is absent and another when the treatment is present. In simplified notation, if a city only completes construction works in the absence of monitoring, we express this in PO notation as C(1) = 0. If there is no monitoring (M = 0) and no completion, then we write C(0) = 0. Conversely, if there is monitoring (M = 1) and completion, we write C(1) = 1, and C(0) = 1 if there is completion without monitoring.

@@

The advantage of this approach is that all assumptions and restrictions imposed to identify the parameters of interest are transparent. It also allows the user to update posterior probabilities based on qualitative data. One reason for this is that with Bayesian inference, we do not need to rely on asymptotic theory to make inferences. Thus, there is no need for a large-n study. Formally speaking, one can make causal queries without any data, based only on prior probabilities for the parameters. Obviously the answer to such causal queries will be vague enough to be almost useless. But the point is that, once you have a prior and a causal model, encoded in a DAG, you can answer causal queries. As one collects data (any data, be it qualitative or quantitative), one can make assumptions about the nature of the data (such as the Data Generating Process), and then update the model and provide new answers to the causal queries, which hopefully will be informative enough to be useful for the task at hand. And even when the model or hypothesis is not completely identified, it can be partially identified [@manski_2003], which is still a useful answer to causal queries.

@@

The evidence gathered is primarily based on documented reports collected by Transparência Brasil and produced by the local OSBs over the course of the project. We also use evidence collected on semi-structured interviews and an online survey with representatives of the 21 CSOs conducted by an independent evaluator of the project. For the quantitative analysis, we use administrative data about construction works data by municipalities and socio-demographic data on municipalities. The questions asked in the survey as well the aggregate answered are presented in annex A.

@@

To put in perspective what was accomplished regarding the completion of all four projects, by the end of the project, of all 135 construction works planned for all 21 municipalities, only 25 were effectively delivered, a 18.5% delivery rate. In Taubaté,  before the monitoring by the local CSO, the delivery rate was zero. After that, the delivery rate was 100% and one of the construction work was a previously paralyzed construction. According to a report by Transparência Brasil [@coelho_etal2021], out of 771 stalled construction projects that were eventually completed, over 50% of them took more than two years to be delivered and 25% took three years or more. As we can see, the odds were against finishing the construction works.

@@

From a causal perspective, we are interested in what is called case-level causal attribution [@humphreys_2023], i.e, did the monitoring cause the delivery of the construction works? In other words, we are conditioning on the fact that M = 1 and C = 1, which constrains the possible types for Taubaté to be $\theta^C_{1,1}$ ou $\theta^C_{0,1}$. Assuming a simple model in which CSO monitoring causes the completion of works (M -> C), and with the eight within-case instances, we do the causal query. Our stimand of interest is the probability that Taubaté is a type $\theta^D_{10}$.

@@

In 2017, the Brazilian municipality of Araucária conducted bidding processes seeking to hire companies to build three nurseries. The project sent technical experts to the 3 planned construction sites, together with volunteers from the local Observatório, to check whether plans were consistent with the physical conditions observed on site. The experts’ assessment was that high-cost contention walls included in the construction plans were unnecessary or could be replaced by low-cost solutions through adjustments. Originally, the contention walls would cost R$ 1,577,338.57 (US$ 419,504.9, at the nominal exchange rate in 2019). The analysis was first directed to the local administration, suggesting changes to the construction plans, which were rejected. The bidding processes went on as initially planned. Subsequently, the experts’ findings were submitted to the Brazilian Supreme Audit Institution, Tribunal de Contas da União (TCU), by Transparência Brasil and Observatório Social do Brazil (see annex 1). TCU recommended that the bidding process should be redone, and construction plans were to be redesigned, with more cost-effective solutions for the contention walls. In the new bidding process, their cost plummeted to R\$ 416,883.17 (USD 110,873.2), a reduction of USD 308,631.7 — or 74% less. To give a sense of the impact on the city budget, the amount saved on these three constructions alone represents about 3% of all capital investments of Araucária in 2018. In short, in a single municipality, the project resulted in savings higher than the cost of the entire project funded by the grantee UNDEF (US$ 220,000.00).

@@

To do so, we collected data on school and nursery construction projects across the entire country, encompassing over two thousand municipalities and more than 14 thousand construction projects, spanning all phases from planning to execution and completion. The @ref(tab:table1) below provides descriptive statistics for these construction projects.

@@

```{r table1, echo=FALSE, message=FALSE, results='asis'}
antes_projeto %>%
  inner_join(depois_projeto, by = "indicator_muni_ot") %>%
  gt() %>%
  tab_header(
    title = "Percentage of finished construction works before the project",
    subtitle = "Situation in August 2017 and August 2019") %>%
   opt_align_table_header(align = "left") %>%
  cols_label(
    indicator_muni_ot = "# of participating cities",
    perc.x = "% (Before Project)",
    perc.y = "% (After Project)"
  ) %>%
    fmt_percent(
    columns = c(perc.x, perc.y),
    decimals = 0
  ) %>%
    tab_source_note(
    source_note = "Construction projects in states SP, MG, SC, PR and RS."
  ) %>%
  tab_options(table.align='left')
```

Our primary objective is to compare the project's effects on the rate of delivery. To estimate the Average Treatment Effect on the Treated (ATT), we employed a matching strategy [@ho_etal2007; @rubin_1973] with a Bayesian logistic regression to create the propensity scores. Balance was successful after matching. We also applied the same procedure to pre-project data to ensure that there were no substantial differences between the treated and control groups.

We estimate all models using the a varying-intercept regression equation [@gelman_2006]:

$$
Y_i = \alpha + \beta_1 T_i + \gamma X_i + \theta Z_i + \epsilon_i
$$

Where $i$ indexes the construction units, $Y_i$ is $1$ if a construction work is finished and $0$ otherwise. $\alpha$ represents the intercept, $\beta$ denotes the average treatment effect on the treated (ATT), and $T_i$ is a binary treatment indicator (whether the municipality is part of Obra Transparent or not). Symbol $\gamma$ is a vector of varying intercepts, $X_i$ is a matrix of random effects for Brazilian municipalities, $\theta$ is a vector of control variables, and $Z_i$ is an array of control variables for case $i$. The error term is represented by $\epsilon_i$. We employed a Bayesian model with rstanarm [@goodrich_etal2023] with default weakly informative priors on the parameters [@gelman_etal2008].

@@

```{r prep data matching,  message=FALSE, echo=FALSE, cache=TRUE}

@@

save(df_balancing_aux1, file = here("Dados", "df_balancing_aux1.RData"))
save(df_balancing_aux_se, file = here("Dados", "df_balancing_aux_se.RData"))
save(df_balancing_aux_post_se, file = here("Dados", "df_balancing_aux_post_se.RData"))
save(data_pos, file = here("Dados", "data_pos.RData"))
```

# these are the no redundancy covariates with and without state covariate
covs.nr <- c('perc_pea18', 'ext_pobres', 'log_rpc', 'p10_ricos_desigualdade', 'freq_escola_15_1', 'analfabetismo',
             'log_renda', 'log_pop', 'percentual_de_execucao', 'tipo_do_projeto')

covs.nr.st <- c(covs.nr, "uf.x")

Table (insert ref number) summarizes the main results of the regression. The first row presents the results for the regression before the project started, where we expect the effect of the project to be zero, since it didn't exist. The second row shows the effect after the end of the project, when we expect the impact to be positive. The first column is an estimate of the posterior mean of the impact of the project and the second one is the posterior standard deviation of the estimate. We also present posterior median and $2.5%$ and $97.5%$ quantiles, in order to quantify the uncertainty around the estimates with a 95% credibility interval.

The hypothesis are corroborated, and we did not find an effect before the starting of the project, but found an effect aftet the end of the project. On average, a city that received the project Obra Transparente increases its delivery rate by almost 10%. These results support our argument and the qualitative evidence we presented that sustained and costly bottom-up accountability improve public policies.

# these are the no redundancy covariates
covs.nr <- c('perc_pea18', 'ext_pobres', 'log_rpc', 'p10_ricos_desigualdade', 'freq_escola_15_1', 'analfabetismo',
             'log_renda', 'log_pop', 'percentual_de_execucao')

ps_spec <- formula(indicator_muni_ot ~ log_rpc  + ext_pobres + freq_escola_15_1 + analfabetismo +
                     log_renda   +  p10_ricos_desigualdade + percentual_de_execucao)

set.seed(1234)
ps_fit_pos <- stan_glm(ps_spec, family=binomial(link='logit'), 
                     data=df_balancing_aux_post_se, algorithm='optimizing', refresh=0, iter = 2000)

# extracting logit pscores
pscores_pre <- apply(posterior_linpred(ps_fit_pre), 2, mean)

# matching
#matching without replacement, original formula
matches_pre <- matching(z=df_balancing_aux_se$indicator_muni_ot, score=pscores_pre, replace=FALSE)
matched_pre <- df_balancing_aux_se[matches_pre$match.ind,]

matches_pos <- matching(z=df_balancing_aux_post_se$indicator_muni_ot, score=pscores_pos, replace=FALSE)
matched_pos <- df_balancing_aux_post_se[matches_pos$match.ind,]

@@

```{r fitting bayesian model, message=FALSE, echo=FALSE, cache = TRUE}

df_balancing_aux_se <- df_balancing_aux_se %>%
  dplyr::select(concluida_num , indicator_muni_ot, code_muni, log_rpc, ext_pobres, freq_escola_15_1,analfabetismo,
                log_renda, p10_ricos_desigualdade, percentual_de_execucao, municipio)

te_spec_nr <- update(ps_spec, concluida_num ~ indicator_muni_ot + (1 | code_muni) + .)

set.seed(120)
reg_ps_pre <- stan_lmer(te_spec_nr, data=df_balancing_aux_se[matches_pre$match.ind,], iter = 4000, refresh=0)

## After
set.seed(120)
reg_ps_pos <- stan_lmer(te_spec_nr, data=df_balancing_aux_post_se[matches_pos$match.ind,], iter = 4000, refresh=0)

@@

```{r processing bayes fit, echo=FALSE,  message=FALSE}
# posterior draws

@@

# Combine summary statistics of posterior simulation draws
b1_df_after_cred <- data.frame(variables = "ATT After project",
                             post_mean = b1_mean_after,
                             post_sd = b1_sd_after,
                             b1_quant_after[2], 
                             b1_quant_after[1],
                             b1_quant_after[3])
b1_df <- bind_rows(b1_df_pre_cred, b1_df_after_cred)
```

```{r table2, echo=FALSE,  message=FALSE, results='asis'}

b1_df %>% gt() %>%
  tab_header(
    title = "Effect of local OSB social monitoring on construction delivery rates",
    subtitle = "Posterior probabilities") %>%
  cols_label(
    variables = "Variables",
    post_mean = "Mean",
    post_sd = "Sd",
    Q50 = "Median",
    Q2.5 = "2.5% CI",
    Q97.5 = "97.5% CI"
  ) %>%
  fmt_number(
    columns = c(post_mean, post_sd, Q50, Q2.5, Q97.5 ),
    decimals = 3,
    use_seps = FALSE
  ) %>%
  tab_source_note(
    source_note = "452 observations for the pre-treatment period and 320 observations for the post-treatment period, resulting from the matching procedure, and encompassing municipalities in the same states as municipalities of Obra Transparente"
  ) %>%
  tab_options(., container.width = 500, container.height =     500)
```

@@
